---
title: "ggplot2 Tips and Tricks and Extensions"
author: "Mike K Smith"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
---

```{r setup}
library(tidyverse)
```

```{r}
plot1 <- mpg %>%
  ggplot(mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

plot1
```

### ggplot2 - Add titles, subtitles, captions, axes labels
```{r}
plot1 +
  labs(title = "Highway mileage is inversely related to engine size",
       subtitle = "Data: mpg",
       x = "Engine size (L)",
       y = "Highway mileage (mpg)",
       caption = "Source: CP1:FI123456")
```


### ggplot2 - Using `%+%` to change the dataset
```{r}
mtcars2 <- mtcars %>%
  rename(hwy = mpg,
         displ = disp)

plot2 <- plot1 %+%
  mtcars2
plot2
```

```{r}
plot3 <- ggplot(data=mpg, mapping=aes(x = class, y = hwy)) +
  geom_boxplot() +
  coord_flip()
plot3

plot4 <- ggplot(mpg) + 
  geom_bar(aes(class,fill=class)) +
  coord_flip()
plot4

plot5 <- mpg %>%
  group_by(class) %>%
  summarise(n = n()) %>%
  mutate(class = fct_reorder(class, n)) %>%
  ggplot(mapping = aes(x = class,fill=class)) + 
  geom_col(mapping = aes(y = n)) +
  coord_flip()
plot5
```


## patchwork - arrange multiple plots

```{r}
library(patchwork)
plot3 + plot4 + plot_annotation(title = "mpg data exploration", 
                                tag_levels="a", tag_suffix = ")")

plot3 + plot4 - plot1 + plot_layout(ncol=1) + plot_annotation(title = "mpg data exploration", 
                                                              tag_levels="i", tag_suffix = ")")
```

## cowplot - templates for publication ready plots
```{r}
library(cowplot)
plot1 + 
  labs(title="Highway mileage is inversely related to engine size", 
       x = "Engine displacement (Litres)",
       y = "Highway mileage (mpg)") +
   theme_cowplot(font_size = 18)

plot1 + 
   theme_minimal_grid()

plot1 + 
   theme_minimal_hgrid()
```

### Adding annotations and text to plots
```{r}
ggdraw(plot1) + 
  draw_label("Draft", color = "light grey", alpha=0.3, size = 100, angle = 45) 
```

### Using `stamp` to label plots
```{r}
stamp(plot1, label = "DRAFT", color = "red")
```

## ggrepel to label points without overlap
```{r}
library(ggrepel)

plot_ugly <- ggplot(mtcars) + 
  aes(
    x = wt, y = mpg,
    label = rownames(mtcars)
  ) + 
  geom_point(color = "red") +
  geom_text()

stamp_ugly(plot_ugly)

ggplot(mtcars) + 
  aes(
    x = wt, y = mpg,
    label = rownames(mtcars)
  ) + 
  geom_point(color = "red") +
  geom_text_repel()
```


## gghighlight only certain observations
```{r}
library(gghighlight)

plot4 + 
  gghighlight(class == "suv")

plot1 + 
  gghighlight(class == "suv") 

plot5 + 
  gghighlight(class == "suv") +
  theme(legend.position="none")
```

## ggridges to compare densities neatly
```{r}
library(ggridges)

ggplot(data = mpg, mapping = aes(x = hwy, y = class)) +
  geom_density_ridges()

## Using fct_reorder from the `forcats` package to sort the factor
## by the median of `hwy` within each class.
ggplot(data = mpg, mapping = aes(x = hwy, y = fct_reorder(class, hwy))) +
  geom_density_ridges()
```

## GGally - Quick exploratory data analysis
```{r}
library(GGally)

mpg %>%
  select(cty, hwy, displ) %>%
  ggscatmat()

mpg %>%
  select(class, cty, hwy, displ) %>%
  ggpairs()
```

## ggforce - additional geoms and extensions to ggplot2
```{r}
library(ggforce)
p <- ggplot(mpg) +
  geom_autopoint() +
  facet_matrix(vars(drv:fl), layer.diag = 2, grid.y.diag = FALSE)
p

p + 
  geom_autodensity(aes(colour = class, fill = class),
                     alpha = 0.4)
```

## ggedit - manipulate plot objects
### Remove plot layers
```{r}
library(ggedit)

## remove loess smooth and replace with a linear regression line
## NB: Uses %>% to pass in the plot object
## Because resulting output is a ggplot2 object we can revert to "+".
plot1 %>%
  ggedit::remove_geom(geom = "smooth") + 
  geom_smooth(method = "lm")
```

### ggedit - interactive editing of plot attributes
```{r, eval = FALSE}
ggedit(plot1)
```

## xgxr - Helper functions for plotting concentration data
```{r}
library(xgxr)

ggplot(data = Theoph, mapping = aes(x = Time, y = conc, group = Subject)) +
  geom_line() + 
  xgx_scale_y_log10()
```

### xgx_stat_ci - stat_summary with additional options for CI type
```{r}
times <- c(0, 0.25, 0.5, 1, 1.5, 2, 3.5, 5, 7, 9, 12, 24)
myTheoph <- Theoph %>%
  mutate(NTPD = metrumrg::snap(Time, times))
table(myTheoph$NTPD)

theoPlot1 <- ggplot(data = myTheoph, aes(x = NTPD,y = conc)) +
  geom_line(mapping = aes(group = Subject), alpha=0.2) + 
  xgx_stat_ci(color = "red") + 
  labs(y = "Concentration (ng/ml)") 
theoPlot1

theoPlot1 +
  xgx_scale_y_log10()
```


## survminer - Plotting survival curves
```{r}
library(survival)
library(survminer)
fit <- survfit(Surv(time, status) ~ sex, data = lung)
ggsurvplot(fit, data = lung)

ggsurvplot(
  fit, 
  data = lung, 
  size = 1,                 # change line size
  palette = 
    c("#E7B800", "#2E9FDF"),# custom color palettes
  conf.int = TRUE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs = 
    c("Male", "Female"),    # Change legend labels
  risk.table.height = 0.25, # Useful to change when you have multiple groups
  ggtheme = theme_bw()      # Change ggplot2 theme
)
```


## Save all ggplot2 objects to working directory

```{r, eval=FALSE}
plotObjects <- Filter( function(x) 'gg' %in% class( get(x) ), ls() )

## Save plot objects individually
sapply(plotObjects, function(x)saveRDS(object = get(x), file = paste0(x,".Rdata",sep="")))

## reload individually
# plot1 <- readRDS("plot1.Rdata")

# Save all plots as one object
# save(list = plotObjects, file = "plots.Rdata")

## test reloading
# remove(list = ls())
# load("plots.Rdata")
```
